---
title: "Report and Code Analyzing Arab Spring"
author: "Nat Larsen"
date: '2023-07-05'
output: github_document
---

``` {r}
# Load libraries
library(readr)
library(ggplot2)
library(dplyr)
library(tidyverse)
```

# Topic Introduction

Starting in late 2010, the Arab Spring was a series of pro-democracy uprisings that swept across several countries in the Middle East and North Africa region. It led to significant political changes, including the overthrow of long-standing authoritarian regimes, calls for democratic reforms, and social and economic upheavals.

# Research Question

To investigate the impact of the Arab Spring the the following years on public opinion, how did the attitudes towards political systems in the Middle East compare across countries? Did citizens prefer democracy or autocracy? Did democratic states have greater satisfaction with their government?

# Description of Dataset

To answer my questions about the public attitudes following the Arab Spring, I downloaded the Wave III data set from [Arab Barometer](https://www.arabbarometer.org/about/the-arab-barometer/) including their 2012-2014 (years directly following Arab Spring) questionnaire data. Arab Barometer is a research organization that conducts public opinion surveys across the Arab World, gaining insight into the social, political, and economic *attitudes* of Arab citizens. Rather than reflecting an objective world, public opinion data reflects the subjective perceptions of people-since I'm interested in citizens' attitudes this data will help with my question, but may limited in providing other insights into the Arab Spring. 

``` {r}
# Importing Data
ABIII_English <- read_csv("data/ABIII_English.csv") # data in `data` folder
```

# Tidying Data

Since I want to focus on the political attitudes of the public surveyed, I want to tidy the data so (1) each variable has its own column, (2) each observation has its own row, and (3) each value must has its own cell. To tidy my dataset, I will combine columns so variables are not. I also want to filter the dataset with the variables I'm interested in. In the original dataset, all the columns are named for the question they correspond to in the questionnaire code (https://www.arabbarometer.org/wp-content/uploads/ABIII_Questionnaire_ENG.pdf), so I should rename the columns from their question number to objects with more meaning. 


```{r}
# Examine dataset to see how to tidy
View(ABIII_English) # looks like meet tidy criteria, but noting that values are different types (strings vs. numbers), and variables have nonsense names

# Select the variables of interest
selected_vars <- c("country", "q5171", "q5172", "q513") # choosing to include the questions about political satisfaction and preferences

# Filter out missing values and subset the data to the selected variables
filtered_data <- ABIII_English %>%
  filter(!is.na(country), !is.na(q5171), !is.na(q5172), !is.na(q513)) %>%
  select(all_of(selected_vars))

# Rename the variables
filtered_data <- filtered_data %>%
  rename(
    country = country,
    pref_demo = q5171, # rename from question code to variable with more meaning
    pref_auth = q5172,
    gov_sat = q513
  )

# View the updated dataframe
head(filtered_data)
```

# Government Satisfaction by Country

Now that my data is tidied and sorted, I first want to investigate my question about the political attitudes in different Middle Eastern countries after the Arab Spring. To do this, I am going to examine government satisfaction across countries by making a barchart with an x-axis of countries, and y-axis of % citizens responded satisfied. 

```{r government satisfaction}
# Create a new dataframe for the graph
graph_gov_sat <- filtered_data %>%
  group_by(country) %>%
  summarize(pct_responses = mean(gov_sat %in% 6:10)) # defining satisfied as responses from 6-10 (indicate more government satisfaction than not)

# Sort the dataframe by the percentage of responses in descending order
graph_gov_sat <- graph_gov_sat[order(graph_gov_sat$pct_responses, decreasing = TRUE), ]

# Create the bar chart
ggplot(graph_gov_sat, aes(x = country, y = pct_responses, fill = country)) +
  geom_bar(stat = "identity") +
  labs(x = "Country", y = "% Responded Satisfied", title = "Satisfaction with Government by Country (2012-2014)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_discrete()
```

The above graph reveals that across the countries, no more than 50% of the population was satisfied with the government—suggesting that everywhere there was not a lot of satisfaction. However, Algeria, Iraq, Jordan, and Kuwait demonstrate the *most* satisfaction, while Libya, Lebanon, and Egypt show the *least* satisfaction. By revealing the countries with the most or least satisfaction, this graph helps visualize which countries experienced which attitude, to help us further investigate reasons why—could there be similarities between the countries with most/least satisfaction? 

# % Citizens Favoring Democracy by Country

To further investigate public attitudes towards political institutions, I want to see what percentage of citizens favor democracy or authoritarianism by plotting countries and the % favor democracy/authoritarianism. 

```{r preference demo/auth}
# Define the columns of interest
columns <- c("pref_demo", "pref_auth")

# Create an empty list to store the results
results <- list()

# Use a for loop to iterate over the columns
for (col in columns) {
  # since both the pref_demo and pref_auth columns use the same criteria to evaluate if citizens         prefer those governments, I can use for loop to manipulate both columns in the same way
  result <- filtered_data %>%
    group_by(country) %>%
    summarize(pct_responses = mean(.data[[col]] %in% c("Very good", "Good"))) # find percentage of "Very good" and "Good" responses for each country
  col_name <- paste0("pct_favor_", col) # rename to represent the specific column processed
  colnames(result)[2] <- col_name
  results[[col]] <- result # add the result to the list
}

# Combine the results into a single dataframe to graph
graph_pref <- Reduce(function(x, y) merge(x, y, by = "country", all = TRUE), results)

# Create the bar chart
ggplot(graph_pref) +
  geom_bar(aes(x = country, y = pct_favor_pref_demo, fill = "Democracy"), stat = "identity", width = 0.5) +
  geom_bar(aes(x = country, y = pct_favor_pref_auth, fill = "Authoritarian"), stat = "identity", width = 0.5) +
  labs(x = "Country", y = "% Responses", title = "Preference for Democracy vs. Authoritarianism (2012-2014)") +
  scale_fill_manual(values = c(Democracy = "blue", Authoritarian = "red"), guide = "none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The above graph shows that across the dataset, all countries have greater preference for democracy (blue) than authoritarianism (red). Algeria, Lebanon, Iraq, and Egypt seem to all have a greater preference for democracy than other countries while Kuwait, Jordan, and Yemen have a greater preference for authoritarianism than other countries. This is an interesting comparison to the graph about government satisfaction—countries may both prefer democracy yet have different levels of government satisfaction (Algeria vs. Egypt). This sparks another question-is there any correlation between government satisfaction and preference for democracy? 

# Government Satisfcation and Democratic Preference 

To answer my question building on government satisfaction and democratic preference, I am going to merge the data frames to make a graph plotting government satisfaction against democratic preference.


```{r satisfaction and democratic preference}
# Merge graph_gov_sat and graph_pref dataframes
graph_sat_pref <- merge(graph_gov_sat, graph_pref, by = "country", all = TRUE)

# Create the scatter plot
ggplot(graph_sat_pref, aes(x = pct_responses, y = pct_favor_pref_demo)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + # line of best fit
  labs(x = "Government Satisfaction", y = "% Favoring Democracy", title = "Government Satisfaction and Preference for Democracy (2012-2014)") +
  theme(plot.title = element_text(hjust = 0.5))
```

The above graph shows that there is some correlation between government satisfaction and % citizens favoring democracy—higher government satisfaction correlates with higher democratic preference. Perhaps in the wake of the Arab Spring and calls for democracy, people felt more satisfied with their governments if they preferred democracy. 

# Summary

This investigation-from finding government satisfaction, to democratic preference, to comparing these two variables against each other-sheds interesting light into the public opinions following the Arab Spring. Overall, I discovered that, at most, half the population was satisfied with the government, meaning that even following the Arab Spring and political changes, government dissatisfaction remained high. However, there remained a large preference for democracy across the Arab World, and those who preferred democracy were more likely to be more satisfied with the government. This suggests that there may be a link between democracy and government satisfaction-or perhaps just a change in political system with government satisfaction. 

While this project sheds light on possible trends between countries, government satisfaction, and democracy, to fully answer my research questions requires further investigation into why citizens might be unsatisfied with their government and what trends between countries may have informed this. I would be interested in advancing these findings through further research with different datasets extending beyond public opinion and perhaps incorporating information such as country democratic status and GDP. It would also be interesting to compare these findings over time, using the other data Arab Barometer generated in later years. Thank you for reading!
